p-adic stability in linear algebra

### 2015-01-18 examples/stability.tex
Small rewriting of Section 3.3 (on characteristic polynomials)

@@

In conclusion, for randomly chosen matrices, approximating $\mathcal{E}$ using the Hodge polygon rarely loses much precision.  However, if the $\sigma_i(M)$ are known to be large or if $M$ is a translate of such a matrix, using lattice precision can be very useful.

>>>>

Write $\charpoly : M_{n,n}(K) \to K[X]$ for the characteristic 
polynomial, and $K_{<n}[X] \subset K[X]$ for the subspace consisting of 
polynomials of degree less than $n$.  Then the differential of 
$\charpoly$ at a point $M$ is
\[
\charp(M) : dM \mapsto \tr(\com(X - M) \cdot dM).
\]
The image is the $K$-span of the entries of $\com(X{-}M)$, which is 
clearly contained within $K_{<n}[X]$.  In fact, the image will equal 
$K_{<n}[X]$ as long as $M$ does not have two Jordan blocks with the same 
generalized eigenvalue. For now on, we make this assumption.

Recall that the \emph{Newton polygon} $\NP(f)$ of a polynomial $f(X) = 
\sum_{i} a_i X^i$ is the lower convex hull of the points $(i, 
\val(a_i))$ and the \emph{Newton polygon} $\NP(M)$ of a matrix $M$ is 
$\NP(\charpoly(M))$.  The \emph{Hodge polygon} of $M$ is the lower 
convex hull of the points $(i, \sum_{j = 1}^{n-i} \sigma_i(M))$. For 
any matrix $M$, the Newton polygon lies above the Hodge polygon 
\cite{kedlaya:padicDiffEq}*{Thm. 4.3.11}.

Such polygons arise naturally in tracking the precision of polynomials 
\cite{caruso-roe-vaccon:14a}*{\S 4.2}.  Any such polygon $P$ yields a 
lattice $\mathcal{L}_P$ in $K_{<n}[X]$ consisting polynomials whose 
Newton polygons lie above $P$.  This lattice is generated by monomials 
$a_iX^i$, where $\val(a_i)$ is the ceiling of the height of $P$ at $i$.  
One uses these polygons to do a certain kind of coordinate-wise 
precision tracking for polynomials, bounding the actual precision 
lattice. We now introduce another polygon that will be bounded by 
between these two, and will provide an estimate for $\mathcal{E}$.

\begin{deftn}
The \emph{precision polygon} $\PP(M)$ of $M$ is the lower convex hull of 
the Newton polygons of the entries of $\com(X{-}M)$.
\end{deftn}

It is clear from the definition $\mathcal L_{\PP(M)} \subset 
\charp(M)(\mathcal{L}_0)$ where $\mathcal{L}_0$ is the standard lattice 
$M_{n,n}{\OK}$. More precisely $\PP(M)$ is the smallest polygon $P$ for 
which the inclusion $\mathcal L_P \subset \charp(M)(\mathcal{L}_0)$. By 
Proposition \ref{prop:precision}, the precision polygon determines the 
best precision loses we can hope if we are encoding precision on 
polynomials by polygons.

It turns out that the precision polygon is related to Hodge and Newton 
polygon. If a polygon $P$ has vertices $(x_i, y_i)$, we let $T_n(P)$ be 
the translated polygon with vertices $(x_i - n, y_i)$.

\begin{prop} \label{prop:polygons}
The precision polygon $\PP(M)$ lies between $T_1(\HP(M))$
and $\NP(M)$.

Moreover $\PP(M)$ and $T_1(\HP(M))$ meet at $0$ and $n{-}1$.
\end{prop}

\begin{proof}
The coefficients of $\charpoly(M)$ can be expressed as traces of 
exterior powers: the coefficient of $X^i$ is $\tr(\Lambda^i(M))$,
which is $\tr(\Lambda^i(U_M) \Lambda^i(\Delta_M) \Lambda^i(V_M))$.  
Computing $\Lambda^i(\Delta_M)$, we get the first statement of
the Proposition. For $i = 1$, we further find that $\PP(M)$ vanishes
at the abscissa $n{-}1$. By definition so does $T_1(\HP(M))$. The fact 
that $\PP(M)$ and $T_1(\HP(M))$ meet at abscissa $0$ follows from
Proposition \ref{prop:detmatrix}.

It remains to prove the comparison with Newton polygon.
Set $f = \charpoly(M)$, set $m_{i,j}$ as the $(i,j)$th entry of $M$, 
$f_{i,j}$ as the $(i,j)$th entry of $\com(X{-}M)$ and $\mu_{i,j} = 
\val(m_{i,j})$. We write $f[k]$ 
for valuation of the coefficient of $X^k$ in $f$, and set $f[-1] = 
+\infty$. The equation $(X{-}M) \cdot \com(X{-}M) = f \cdot I$ yields, 
for all $i$ and $k$,
\begin{align*}
f[k] &\ge \inf(f_{i,i}[k{-}1], \mu_{i,0} + f_{0,i}[k], \ldots, 
\mu_{i,n} + f_{n,i}[k]) \\
&\ge \inf(f_{i,i}[k{-}1], f_{j,i}[k]),
\end{align*}
with the infimum is taken over $j$. 
Taking lower convex hulls and noting that $\PP(M)$ is nonincreasing
(it follows from the comparison with the Hodge polygon), we get the 
result.
\end{proof}

\begin{rem}
Experiments actually support the following stronger result: $\PP(M)$ is 
bounded above by $T_1(\HP(M))$.
\end{rem}

For many matrices, $\PP(M) = T_1(\HP(M))$.  For random matrices over 
$\Z_2$, the $2$-adic precision polygon is equal to the Hodge polygon in 
$99.5\%$ of cases in dimension $4$, down to $99.1\%$ in dimension $8$.  
Over $\Z_3$, the fraction rises to $99.98\%$, with no clear dependence on 
dimension.
Empirically, $\PP(M)$ seems most likely to differ from $T_1(\HP(M))$ at 
$1$, corresponding to the precision of the linear term of the 
characteristic polynomial.

Of course, $\PP(M)$ does necessarily agree with the precision lattice 
$\mathcal{E} = \charp(M)(\mathcal L_0)$.  In $11\%$ of cases in 
dimension $3$, up to $15\%$ of cases in dimension $8$, the index of 
$\mathcal{E}$ in $\mathcal{L}_{\PP(M)}$ is greater than $1$.  This 
percentage increases as $\val(\det(M))$ increases, reaching $34\%$ in 
dimension $9$ for matrices constrained to have determinant with $2$-adic 
valuation $12$.
Moreover, one can construct examples with $[\mathcal{L}_{\PP(M)} : 
\mathcal{E}]$ arbitrarily large.  Suppose the $\sigma_i(M)$ are large.  
Proposition \ref{prop:polygons} implies that $\mathcal{E}$ is 
contained within $\O_{K, <n}[X]$ with index at least $\sum_{i=1}^{n-1} 
\sigma_i(M)$.  The precision lattice of $1 + M$ is obtained from 
$\mathcal{E}$ via the transformation $X \mapsto 1 + X$, but $\PP(1 + M)$ 
is now flat with height $0$.

In conclusion, for randomly chosen matrices, approximating $\mathcal{E}$ 
using the Hodge polygon loses precision but not so much. Nonetheless, as 
already illustrated for matrix multiplication, this loses of precision 
may accumulate and become significant if we need to compute many 
characteristic polynomials. In the case of a single characteristic 
polynomial, we note moreover that if the $\sigma_i(M)$'s are known to be 
large or if $M$ is a translate of such a matrix, using lattice precision 
can be very useful.

\begin{comment}
\subsection{Characteristic polynomials} \label{ssec:charpoly}

@@

In conclusion, for randomly chosen matrices, approximating $\mathcal{E}$ using the Hodge polygon rarely loses much precision.  However, if the $\sigma_i(M)$ are known to be large or if $M$ is a translate of such a matrix, using lattice precision can be very useful.
\end{comment}
